---
title: "Covid19_Report"
author: "Student"
date: "2024-09-06"
output:
  pdf_document: default
  html_document: default
---
## Introduction
The COVID-19 pandemic, has significantly affected public health, economies, and daily life around the world. In this project we will analyze COVID-19 data to better understand the pandemic's dynamics and impact. By systematically examining and interpreting COVID-19 data, this project aims to enhance our understanding, improve response strategies, and ultimately contribute to controlling the pandemic.

In this project we will import, tidy and analyze the COVID19 data set from the Johns Hopkins github site. This data set contains daily time series summary tables, including confirmed, deaths and recovered. All data is read in from the daily case report. 

Source: "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
(https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)
We will be using "time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv" and  "time_series_covid19_deaths_global.csv" files for this project.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(leaflet)
```

## Data Collection
We will start by reading in the data from the four main csv files.
```{r}
## Get current data in the four files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c( "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv",
"time_series_covid19_confirmed_US.csv",  
"time_series_covid19_deaths_US.csv" 
)
urls <- str_c(url_in, file_names)
urls
```

Let us read in the data and see what we have.
```{r}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])
```
Let us have a look at the few rows of global_cases.
```{r}
head(global_cases)
```

## Data Cleaning and Preparation
Next, we would like to tidy these data sets and put each variable (date, cases, deaths) in their own column. Also, we do not need lat and long for the analysis that we are planning so we will remove them and rename Region and State to be more R friendly.
```{r}
global_cases <- global_cases %>%
  pivot_longer(cols=-c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))
global_cases
```

```{r}
global_deaths <- global_deaths %>%
  pivot_longer(cols=-c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))
global_deaths
```
Now let us combine cases and deaths into one variable called global.
```{r}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`)%>%
  mutate(date = mdy(date))
global
```
Now let us have a look at the summary of the data.
```{r}
summary(global)
```
Next, let us filter out rows where there are 0 cases.
```{r}
global<- global %>% filter(cases>0)
summary(global)
```
Let us check if the maximum value is valid.
```{r}
global %>% filter(cases>103000000)
```
Let us have look at US cases.
```{r}
US_cases
```
We will pivot the cases.
```{r}
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
                          names_to = "date",
                          values_to = "cases")
US_cases
```
Next, we will select columns from Admin2 to cases and change date from char to date object.
```{r}
US_cases <- US_cases %>% select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
US_cases
```
Let us have look at US deaths.
```{r}
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
                          names_to = "date",
                          values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
US_deaths
```
Let us join US_cases and US_deaths into one US
```{r}
US <- US_cases %>%
  full_join(US_deaths, by = c("Admin2","Province_State","Country_Region","Combined_Key","date"))
US
```
We do not have Population data for gobal data, so we will add it to our global dataset.
```{r}
# Creating Combined Key
global <- global %>%
  unite("Combined_Key", c(Province_State,Country_Region),
        sep= ", ",
na.rm = TRUE,
remove = FALSE)
global
```
```{r}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```
Let us join it to global.
```{r}
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
global
```

## Exploratory Data Analysis (EDA)
Let us look at the summary of the data.
```{r}
## Summary
summary(global)
```

### Visualizing the number of cases for each state on each date
```{r}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_by_state 
```

### Visualizing the total for the US
```{r}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
US_totals
```
Let us look at the end of this.
```{r}
tail(US_totals)
```

#### Visualization of total number of cases and deaths in the United States.
```{r}
US_totals %>%
  ggplot(aes(x = date, y = cases))+
geom_line(aes(color="cases"))+
geom_point(aes(color="cases"))+
geom_line(aes(y= deaths, color="deaths"))+
geom_point(aes(y= deaths, color="deaths"))+
scale_y_log10()+
theme(legend.position = "bottom",
      axis.text.x = element_text(angle = 90))+
labs(title = "Covid19 in US", y = NULL)

```

### Visualization of total number of cases and deaths in New York.
```{r}
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases))+
geom_line(aes(color="cases"))+
geom_point(aes(color="cases"))+
geom_line(aes(y= deaths, color="deaths"))+
geom_point(aes(y= deaths, color="deaths"))+
scale_y_log10()+
theme(legend.position = "bottom",
      axis.text.x = element_text(angle = 90))+
labs(title = str_c("Covid19 in ",state), y = NULL)

```

### Visualization of total number of cases and deaths in Pennsylvania.
```{r}
state <- "Pennsylvania"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases))+
geom_line(aes(color="cases"))+
geom_point(aes(color="cases"))+
geom_line(aes(y= deaths, color="deaths"))+
geom_point(aes(y= deaths, color="deaths"))+
scale_y_log10()+
theme(legend.position = "bottom",
      axis.text.x = element_text(angle = 90))+
labs(title = str_c("Covid19 in ",state), y = NULL)

```

Find the date up to which we have data.
```{r}
max(US_totals$date)
```
Find the maximum number of deaths.
```{r}
max(US_totals$deaths)
```

### States with maximum and minimum cases

Next, let us see the states with maximum and minimum cases.
```{r}
# Determine the states with maximum and minimum cases
states_cases_summary <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(total_cases = sum(cases), .groups = 'drop')

state_max_cases <- states_cases_summary %>%
  filter(total_cases == max(total_cases)) %>%
  pull(Province_State)

state_min_cases <- states_cases_summary %>%
  filter(total_cases == min(total_cases)) %>%
  pull(Province_State)

```

We will then filter the data for these states, so that we can visualize them.
```{r}
# Filter data for these states
states_to_plot <- c(state_max_cases, state_min_cases)
plot_data <- US_by_state %>%
  filter(Province_State %in% states_to_plot) %>%
  filter(cases > 0)  # Filter out rows where cases are 0

# Plot data for states with max and min cases
ggplot(data = plot_data, aes(x = date)) +
  geom_line(aes(y = cases, color = "cases")) +
  geom_point(aes(y = cases, color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  scale_color_manual(values = c("cases" = "royalblue", "deaths" = "salmon")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state_max_cases, " and ", state_min_cases),
       y = "Count (Log Scale)",
       color = "Legend") +
  facet_wrap(~ Province_State)
```

In the plot above we see that we get State with the Maximum number of cases as California and with minimum number of cases as "Diamond Princess" but this is not the name of any state in the US. In fact, it is the name of the cruise ship that had a notable COVID-19 outbreak. So, let us check the unique values in the Province_State column.
```{r}
# Check unique values in the Province_State column
unique_states <- US_by_state %>%
  distinct(Province_State) %>%
  pull(Province_State)

# Print unique states to the console
print(unique_states)
```
We see that "Diamond Princess" and "Grand Princess" are cruise ships and not US states or U.S. territories.
```{r}
# Exclude non-state entities
states_cases_summary <- US_by_state %>%
  filter(!Province_State %in% c("Diamond Princess", "Grand Princess", "Other non-state entities")) %>%
  group_by(Province_State) %>%
  summarize(total_cases = sum(cases), .groups = 'drop')

# Proceed with identifying the states with max and min cases
state_max_cases <- states_cases_summary %>%
  filter(total_cases == max(total_cases)) %>%
  pull(Province_State)

state_min_cases <- states_cases_summary %>%
  filter(total_cases == min(total_cases)) %>%
  pull(Province_State)

# Filter data for these states
states_to_plot <- c(state_max_cases, state_min_cases)

plot_data <- US_by_state %>%
  filter(Province_State %in% states_to_plot) %>%
  filter(cases > 0)  # Filter out rows where cases are 0

# Plot data for states with max and min cases
ggplot(data = plot_data, aes(x = date)) +
  geom_line(aes(y = cases, color = "cases")) +
  geom_point(aes(y = cases, color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  scale_color_manual(values = c("cases" = "royalblue", "deaths" = "salmon")) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID-19 in ", state_max_cases, " and ", state_min_cases),
       y = "Count (Log Scale)",
       color = "Legend") +
  facet_wrap(~ Province_State)

```

From the above plot we see that American Samoa (a U.S. territory), has the minimum number of cases and the State of California has the maximum number of cases.

## Data Modeling

Now we will build a linear regression model to predict deaths per million. We will plot this predicted_deaths_per_mill with actual deaths_per_mill to evaluate the performance of our regression model. 
```{r}
# Remove rows with NA, NaN, or Inf values
cleaned_data <- US_by_state %>%
  filter(!is.na(deaths_per_mill) & !is.nan(deaths_per_mill) & !is.infinite(deaths_per_mill) &
         !is.na(cases) & !is.nan(cases) & !is.infinite(cases) &
         !is.na(Population) & !is.nan(Population) & !is.infinite(Population))
# Fit a linear regression model
model <- lm(deaths_per_mill ~ cases + Population, data = cleaned_data)

# Summary of the model
summary(model)
```

```{r}
# Predict deaths per million
cleaned_data$predicted_deaths_per_mill <- predict(model, newdata = cleaned_data)

# Plot actual vs predicted
ggplot(cleaned_data, aes(x = deaths_per_mill, y = predicted_deaths_per_mill)) +
  geom_point(color = "cyan", size = 3, alpha = 0.6) +
  geom_abline(intercept = 0, linewidth = 3, slope = 1, color = "red") +
  labs(x = "Actual Deaths per Million", y = "Predicted Deaths per Million",title = "Actual vs Predicted Deaths per Million")+
theme_minimal() 
```

From the above plot we see that the majority of points are close to the red line, it indicates that the model’s predictions the model’s predictions are generally accurate, suggesting good model performance.

## Results and Interpretation
* Cases have stabilized due to widespread vaccination, natural immunity, and public health measures.
* We have the data till 2023-03-09.
* Maximum maximum number of deaths in US is 1123836.
* American Samoa (a U.S. territory), has the minimum number of cases and the State of California has the maximum number of cases.

## Conclusion and Bias

COVID-19 data analysis is a critical for navigating the complexities of the pandemic. By leveraging data, researchers, policymakers, and health professionals can better understand the dynamics of the virus, implement effective interventions, and ultimately work towards controlling the pandemic and mitigating its effects on society.

Bias: There may be inconsistencies in how cases and deaths are recorded and reported, because different states may have different reporting practices. Some cases or deaths may not be reported. Unequal access to testing across states and communities can result in some areas being under-tested, affecting reported case numbers.

